# Links to Book Code and Additional Data {#book-data}

The code and data and an RStudio project can be downloaded with 

```{r}
#| code-fold: false
#| eval: false
usethis::use_zip(url="https://dicook.github.io/mulgar_book/code_and_data.zip")
```

Alternatively, individual files can be downloaded from the links below.

## Additional data

`r ifelse(knitr::is_html_output(), '@tbl-datalinks-html', '@tbl-datalinks-pdf')` lists additional data available on the book web site at https://dicook.github.io/mulgar_book/data. 

```{r}
#| message: false
#| echo: false
#| warning: false
library(readr)
book_data <- read_csv("misc/book_data.csv") 
```

::: {.content-visible when-format="html"}

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: tbl-datalinks-html
#| tbl-cap: "Links to other data sets used on the book."
library(gt)
library(gtExtras)
library(gtsummary)
book_data |> 
  dplyr::select(Description, Link) |>
  gt() |>
  cols_width(Description~px(200), 
             Link~px(400))
```
:::

::: {.content-visible when-format="pdf"}
```{r}
#| message: false
#| echo: false
#| warning: false
#| label: tbl-datalinks-pdf
#| tbl-cap: "Links to other data sets used on the book."
library(kableExtra)
book_data |> 
  dplyr::select(Filename, Description, Chapter) |>
  knitr::kable(format="latex", booktabs = T) |>
  kable_paper(full_width = F) |>
  kable_styling() |>
  column_spec(1, width = "4cm", monospace=TRUE) |>
  column_spec(2, width = "5cm") |>
  column_spec(3, width = "2cm")
```
:::

## Code files

`r ifelse(knitr::is_html_output(), '@tbl-codelinks-html', '@tbl-codelinks-pdf')` lists additional data available on the book web site at https://dicook.github.io/mulgar_book/code. 

```{r}
#| message: false
#| echo: false
book_code <- read_csv("misc/book_code.csv") 
```

::: {.content-visible when-format="html"}

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: tbl-codelinks-html
#| tbl-cap: "Links to code used on the book."
book_code %>% 
  mutate(Link = purrr::map(Link, gt::html)) %>% 
  gt() %>% 
  cols_width(Chapter~px(300), Link~px(300)) %>%
  cols_align( align = c("left"), columns = everything())
```

:::

::: {.content-visible when-format="pdf"}
```{r}
#| message: false
#| echo: false
#| warning: false
#| label: tbl-codelinks-pdf
#| tbl-cap: "Links to code used on the book."
library(kableExtra)
book_code |> 
  dplyr::select(Filename, Chapter) |>
  knitr::kable(format="latex", booktabs = T) |>
  kable_paper(full_width = F) |>
  kable_styling() |>
  column_spec(1, width = "4cm", monospace=TRUE) |>
  column_spec(2, width = "5cm") 
```
:::
