# Preface {#preface .unnumbered}

<!-- Why write this book

-   an introduction to modern statistical methods through the lens of visualisation
-   focus on data analysis and computing over mathematics
-   intuit high dimensional data through dynamic graphics
-   use R + graphics as tool for augmenting understanding of data, models and inferences(?)
-->

Although there are many resources available for data visualization, there are few comprehensive resources on high-dimensional data visualisation. This book fills this gap by providing a comprehensive and up-to-date guide to visualising high-dimensional data and models, with R.

High-dimensional data spaces are fascinating places. You may think that there's a lot of ways to plot one or two variables, and a lot of types of patterns that can be found. You might use a density plot and see skewness or a dot plot to find outliers. A scatterplot of two variables might reveal a non-linear relationship or a barrier beyond which no observations exist. We don't as yet have so many different choices of plot types for high-dimensions, but these types of patterns are also what we seek in scatterplots of high-dimensional data. The additional dimensions can clarify these patterns, that clusters are likely to be more distinct. Observations that did not appear to be very different can be seen to be lonely anomalies in high-dimensions, that no other observations have quite the same combination of values. 

## What's in this book? {-}

The book is divided into these parts:

- **Introduction**: Here we introduce you to high-dimensional spaces, how they can be visualised, and notation that is useful for describing methods in later chapters.   
- **Dimension reduction**: This part covers linear and non-linear dimension reduction. It includes ways to help decide on the number of dimensions needed to summarise the high dimensional data, whether linear dimension reduction is appropriate, detecting problems that might affect the dimension reduction, and examining how well or badly a non-linear dimension reduction is representing the data. 
- **Cluster analysis**: This part described methods for finding groups in data. Although it includes an explanation of a purely graphical approach, it is mostly on using graphics in association with numerical clustering algorithms. There are explanations of assessing the suitability of different numerical techniques for extracting clusters, based on the data shapes, evaluating the clustering result, and showing the solutions in high dimensions.
- **Classification**: This part describes methods for exploring known groups in the data. You'll learn how to check model assumptions, to help decide if a method is suited to the data, examine classification boundaries and explore where errors arise. 
- **Miscellaneous**: The material in this part focuses on examining data from different contexts. This includes multiple time series, longitudinal data. A key pre-processing step is to convert the data into Euclidean space. 

In each of these parts an emphasis is also showing your model with your data in the high dimensional space. 

Our hopes are that you will come away with understanding the importance of plotting your high dimensional data as a regular step in your statistical or machine learning analyses. There are many examples of what you might miss if you don't plot the data. Effective use of graphics goes hand-in-hand with analytical techniques. With high dimensions visualisation is a challenge but it is fascinating, and leads to many surprising moments.

## Audience {-}

High-dimensional data arises in many fields such as biology, social sciences, finance, and more. Anyone who is doing exploratory data analysis and model fitting for more than two variables will benefit from learning how to effectively visualise high-dimensions. This book will be useful for students and teachers of mulitvariate data analysis and machine learning, and researchers, data analysts, and industry professionals who work in these areas. 

## How to use the book? {.unnumbered}

The book is written with explanations followed by examples with R code. The toolbox chapter provides an overview of the primary high-dimensional visualisation methods. The remaining chapters focus on different application areas and how to use the high-dimensional visualisation to complement commonly used analytical methods. 

## What should I know before reading this book? {.unnumbered}

The examples assume that you already use R, and have a working knowledge of base R and tidyverse way of thinking about data analysis. It also assumes that you have some knowledge of statistical methods, and some experience with machine learning methods. 

If you feel like you need build up your skills in these areas in preparation for working through this book, these are our recommended resources:

- [R for Data Science](https://r4ds.had.co.nz) by Wickham and Grolemund for learning about data wrangling and visualisation. 
- [Introduction to Modern Statistics](https://openintro-ims.netlify.app) by 
Çetinkaya-Rundel and Hardin to learn about introductory statistics.
- [Hands-On Machine Learning with R](https://bradleyboehmke.github.io/HOML/) by Boehmke and Greenwell to learn about machine learning.

## Setting up your workflow {.unnumbered}

To get started set up your computer with the current versions of [R](https://cran.r-project.org) and [Rstudio Desktop](https://posit.co/download/rstudio-desktop/).

In addition, we have made an R package to share the data and functions used in this book, called [`mulgar`](http://dicook.github.io/mulgar).[^mulga][^photo]

[^mulga]: Mulga is a type of Australian habitat composed of woodland or open forest dominated by the mulga tree. Massive clearing of mulga led to the vast wheat fields of Western Australia. Here **mulgar** is an acronym for **MUL**tivariate **G**raphical **A**nalysis with **R**.

[^photo]: Photo of mulga tree taken by L. G. Cook.

```{r}
#| label: setup
#| eval: false
#| code-fold: false
install.packages("mulgar", dependencies=TRUE)
#| or the development version
devtools::install_github("dicook/mulgar")
```

```{r}
#| include: false
#| echo: false
# automatically create a bib database for R packages
knitr::write_bib(c(
 "aweSOM", "classifly", "colorspace", "crosstalk",
 "detourr",
 "dplyr",
 "e1071", "geozoo", "GGally", "ggdendro", 
 "ggplot2", "ggpubr", "gridExtra", "gt", 
 "gtExtras", "ggrepel",
 "gtsummary", "ggthemes", "htmlwidgets", "kableExtra", 
 "kohonen", "langevitour", "liminal", "MASS", "mclust",
  "mulgar", "palmerpenguins", "patchwork", 
 "plotly", 
 "randomForest", "readr", "rpart", "rpart.plot",
  "Rtsne", "tourr"), 'packages.bib')
```

## How do I report an  error?

If you encounter an error, you can report it as an issue at the [Github repo for this book](https://github.com/dicook/mulgar_book). 

Please make a small [reproducible example](https://reprex.tidyverse.org) and report the error encountered. Reproducible examples have these components:

- a small amount of data 
- small amount of code that generates the error
- copy of the error message that was generated

```{r}
#| eval: false
#| echo: false
# Colour testing
hcl.colors(12, palette="Zissou 1")
pie(rep(1, 12), col = hcl.colors(12, "Zissou 1"), main = "HCL")
```

## License

The online version of this book is licensed under a
[Creative Commons Attribution-NonCommercial-ShareAlike 4.0 International License](https://creativecommons.org/licenses/by-nc-sa/2.0/).

[![](https://upload.wikimedia.org/wikipedia/commons/1/12/Cc-by-nc-sa_icon.svg)](https://creativecommons.org/licenses/by-nc-sa/2.0/)
