# Preface {#preface .unnumbered}

<!-- Why write this book

-   an introduction to modern statistical methods through the lens of visualisation
-   focus on data analysis and computing over mathematics
-   intuit high dimensional data through dynamic graphics
-   use R + graphics as tool for augmenting understanding of data, models and inferences(?)
-->

Although there are many resources available for data visualization, there are few comprehensive resources on high-dimensional data visualisation. This book fills this gap by providing a comprehensive and up-to-date guide to visualising high-dimensional data and models, with R.

High-dimensional data spaces are fascinating places. You may think that there's a lot of ways to plot one or two variables, and a lot of types of patterns that can be found. You might use a density plot and see skewness or a dot plot to find outliers. A scatterplot of two variables might reveal a non-linear relationship or a barrier beyond which no observations exist. We don't as yet have so many different choices of plot types for high-dimensions, but these types of patterns are also what we seek in scatterplots of high-dimensional data. The additional dimensions can clarify these patterns, that clusters are likely to be more distinct. Observations that did not appear to be very different can be seen to be lonely anomalies in high-dimensions, that no other observations have quite the same combination of values. 

## Audience {.unnumbered}

High-dimensional data arises in many fields such as biology, social sciences, finance, and more. Anyone who is doing exploratory data analysis and model fitting for more than two variables will benefit from learning how to effectively visualise high-dimensions. This book will be useful for students and teachers of mulitvariate data analysis and machine learning, and researchers, data analysts, and industry professionals who work in these areas. 

## How to use the book? {.unnumbered}

The book is written with explanations followed by examples with R code. The toolbox chapter provides an overview of the primary high-dimensional visualisation methods. The remaining chapters focus on different application areas and how to use the high-dimensional visualisation to complement commonly used analytical methods. 

## What should I know before reading this book? {.unnumbered}

The examples assume that you already use R, and have a working knowledge of base R and tidyverse way of thinking about data analysis. It also assumes that you have some knowledge of statistical methods, and some experience with machine learning methods. 

If you feel like you need build up your skills in these areas in preparation for working through this book, these are our recommended resources:

- [R for Data Science](https://r4ds.had.co.nz) by Wickham and Grolemund for learning about tidyverse. 
- [Introduction to Modern Statistics](https://openintro-ims.netlify.app) by 
Çetinkaya-Rundel and Hardin to learn about introductory statistics.
- [Hands-On Machine Learning with R](https://bradleyboehmke.github.io/HOML/) by Boehmke and Greenwell to learn about machine learning.

## Setting up your workflow {.unnumbered}

To get started set up your computer with the current versions of [R](https://cran.r-project.org) and [Rstudio Desktop](https://posit.co/download/rstudio-desktop/).

In addition, we have made an R package to share the data and functions used in this book, called [`mulgar`](http://dicook.github.io/mulgar).[^mulga][^photo]

[^mulga]: Mulga is a type of Australian habitat composed of woodland or open forest dominated by the mulga tree. Massive clearing of mulga led to the vast wheat fields of Western Australia. Here **mulgar** is an acronym for **MUL**tivariate **G**raphical **A**nalysis with **R**.

[^photo]: Photo of mulga tree taken by L. G. Cook.

```{r}
#| label: setup
#| eval: false
#| code-fold: false
install.packages("mulgar")
#| or the development version
devtools::install_github("dicook/mulgar")
```

```{r}
#| include: false
#| echo: false
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages()), 'packages.bib')
```

## How do I report an  error?

If you encounter an error, you can report it as an issue at the [Github repo for this book](https://github.com/dicook/mulgar_book). 

Please make a small [reproducible example](https://reprex.tidyverse.org) and report the error encountered. Reproducible examples have these components:

- a small amount of data 
- small amount of code that generates the error
- copy of the error message that was generated

```{r}
#| eval: false
#| echo: false
# Colour testing
hcl.colors(12, palette="Zissou 1")
pie(rep(1, 12), col = hcl.colors(12, "Zissou 1"), main = "HCL")
```
