--- 
title: "Interactive and dynamic graphics for multivariate data using R"
author: "Di Cook, Ursula Laa, Stuart Lee, Earo Wang"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
description: "This is a book about exploring multivariate data."
---

# Preface {-}

Why write this book

* an introduction to modern statistical methods through the lens of visualisation
* focus on data analysis and computing over mathematics
* intuit high dimensional data through dynamic graphics
* use R + graphics as tool for augmenting understanding of data, models and inferences(?)

## Audience


* researchers/students who need to analyse multivariate data
* data scientists looking to expand their toolkit and not just run to a model straight away


## How to use the book?

* How can does the material fit together  - some diagrams


### What do we assume about you?

* already knows a bit of R and a bit of stats already (?)
* expect you to have work through the code examples
* be actively involved with the text through excercies and examples


## Setting up your workflow

* Setting up R and Rstudio
* What packages to install
* tidyverse style?

```{r eval=FALSE}
install.packages("mulgar")
# or the development version
# devtools::install_github("dicook/mulgar")
```


```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```
