# Preface {#preface .unnumbered}

Why write this book

-   an introduction to modern statistical methods through the lens of visualisation
-   focus on data analysis and computing over mathematics
-   intuit high dimensional data through dynamic graphics
-   use R + graphics as tool for augmenting understanding of data, models and inferences(?)

## Audience {.unnumbered}

-   researchers/students who need to analyse multivariate data
-   data scientists looking to expand their approaches and not just run to a model straight away

## How to use the book? {.unnumbered}

-   How does the material fit together - some diagrams

### What do we assume about you? {.unnumbered}

-   already knows a bit of R and a bit of stats already (?)
-   expect you to have work through the code examples
-   be actively involved with the text through excercies and examples

## Setting up your workflow {.unnumbered}

-   Setting up R and Rstudio
-   What packages to install
-   tidyverse style?

```{r eval=FALSE}
install.packages("mulgar")
# or the development version
# devtools::install_github("dicook/mulgar")
```

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```
